---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading and preprocessing the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(ggplot2)


##download and read csv into 'data'
temp <- tempfile()
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, temp)
df <- read.csv(unz(temp,"activity.csv"), stringsAsFactors = FALSE)
unlink(temp)
df.missing <- df

```

## What is mean total number of steps taken per day?

```{r mean_total_steps}
hist(df$steps)
```

The mean total number of steps is **`r mean(df$steps, na.rm = TRUE)`**.
The median total number of steps is **`r median(df$steps, na.rm = TRUE)`**.

## What is the average daily activity pattern?

```{r average_pattern}

df$steps[is.na(df$steps)] <- 0

df.ts <- df %>%
  group_by(interval) %>%
  summarize(total = mean(steps))

ggplot(df.ts, aes(x = interval, y = total, group = 1)) +
  geom_line()
```

The interval with the largest average number of steps is **`r df.ts$interval[df.ts$total == max(df.ts$total)]`**.

## Imputing missing values

```{r missing values}
na.count <- df.missing %>%
  summarize_each(funs(sum(is.na(.))))
print(na.count) 

##replace NA with mean for that interval
df.missing <- df.missing %>%
  group_by(interval) %>%
  mutate(steps.na = ifelse(is.na(steps), mean(steps, na.rm =TRUE), steps))

hist(df.missing$steps.na)
```

The mean total number of steps is **`r mean(df.missing$steps.na, na.rm = TRUE)`**.
The median total number of steps is **`r median(df.missing$steps.na, na.rm = TRUE)`**.
It doesn't appear that imputing NAs with the mean changes the average steps but did increased the frequency counts as expected.

## Are there differences in activity patterns between weekdays and weekends?

```{r patterns}
##create factor for weekday or weekend
df.missing$weekend <- as.factor(
  ifelse(weekdays(as.Date(df.missing$date, "%Y-%m-%d")) %in% c("Saturday", "Sunday"),"weekend","weekday"))

## create plot comparing steps on weekday versus weekend
df.week <- df.missing %>%
  group_by(weekend, interval) %>%
  summarise(total = sum(steps.na))

ggplot(df.week, aes(x = interval, y = total, group = 1)) +
  geom_line() +
  facet_wrap(~weekend, ncol = 1)
